Synthesizing Symmetric Lenses

### 2018-07-01 papers/SymmetricLenses/main.tex
edits in progress to section 2

Given a lens $\Lens$ between two stochastic regular expressions $\Regex$ and
$\RegexAlt$, we can calculate the expected number of bits required
to recover a string $s$ in $\LanguageOf{\Regex}$ from a synchronized
string $t$ in $\LanguageOf{\RegexAlt}$. 
This expected value is just the conditional entropy of the $\Regex$ given
$\RegexAlt$ and $\Lens$. 
To ensure symmetry, we seek to find the lens that
minimizes the sum of that entropy with the conditional entropy of $\RegexAlt$
given $\Regex$ and $\Lens$.

\ksf{This piece might better go somewhere deeper in the paper.}
Our algorithm guesses a reasonable probability distribution from
the syntax, though if users manually provided stochastic regular expressions, or
a large suite of data in the formats through which we could learn the annotations,
we could instead use that.

Now that we have a notion of best, we need a way to search for that best lens.
Our overall approach involves two communicating synthesizers: one (\RXSearch)
that proposes regular expression pairs, semantically but not syntactically
equivalent to the originals; one (\GreedySynth) that tries to find the best lens
within a class of lenses, characterized by such regular expression pairs. Given
two regular expressions, \GreedySynth can quickly search through the class of
lenses that can be typed between them, where all instances of type equivalence
involve no star equivalences. For example, consider the lens shown in
Figure~\ref{fig:example_lens}. This lens would \emph{not} be in the class
characterized by \lstinline{emp_salaries} and \lstinline{emp_insurance}, as that
typing derivation requires unrolling \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into \lstinline{"" | (("\n" . emp_salary) . ("\n" . emp_salary)$^*$)}.

How does \RXSearch propose stochastic regular expression pairs? We define
rewrite rules for stochastic regular expressions, and prove those rewrites
preserve language and probability distributions. By applying these rewrites, we
can generate new regular expression pairs, syntactically distinct but
semantically equivalent to the originals. \RXSearch proposes the regular
expression pairs which are most similar to the originals first (in that they
require fewer rewrite rules to generate), and iteratively propose pairs that are
increasingly dissimilar to the original pair (in that they require more rewrite
rules to generate).

Critically for this work, \GreedySynth efficiently searches for one of the best
lenses in a class of lenses, characterized by a stochastic regular expression
pair. \GreedySynth works by searching through a smaller search space. First,
\GreedySynth converts the provided stochastic regular expression pair into a
\emph{stochastic DNF regular expression} pair. Then, the algorithm uses the
syntax of the DNF regular expression pair to greedily find a highly normalized
version of a simple symmetric lens, a \emph{symmetric DNF lens}. Finally, this
symmetric DNF lens is converted into a simple symmetric lens. DNF Lenses include
neither a composition operator nor a type equivalence rule. Unlike simple
symmetric lenses, for which a single lens term is typed between many possible
regular expression pairs, each DNF lens is typed only between a single regular
expression pair. The relationship between DNF lens syntax and DNF regular
expression syntax is very tight, permitting an efficient
synthesis algorithm for DNF lenses from DNF regular expressions.

>>>>

This infrastructure gives us what we need to find the best
satisfying lens from a collection of satisfying lenses.  All such
lenses must synchronize data between the same two 
stochastic regular expressions $\Regex$ and $\RegexAlt$.
For any such lens $\Lens$, we can calculate the expected number of
bits required 
to recover a string $s$ in $\LanguageOf{\Regex}$ from a synchronized
string $t$ in $\LanguageOf{\RegexAlt}$. 
This expected number of bits is just the conditional entropy of
$\Regex$ given $\RegexAlt$ and $\Lens$. 
To ensure symmetry, the entropy we assign to $\Lens$ is the sum of 
the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ with 
the conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$.

To obtain stochastic regular expressions from the plain regular
expressions that we expect users to supply, we use a heurstic 
that attempts to assign
probability annotations that give each string in the language equal
probablity.  We can also take as input a stochastic regular
expression rather than a plain one, which allows the user to
specify a data-specific probability distribution.  Alternatively,
the user could provide a large suite of example data from which we
could infer the necessary annotations to model the supplied data.

\ksf{Should we be talking about ``best'' or just ``good'' or ``low''?
There are typically a collection of ``best'' lenses, and we choose the
one that requires the simplest permutation.}
Now that we have a notion of best, we need a way to search for that
best lens.  Previous work on synthesis~\ksf{add citations} has shown
that we can use types to dramatically restrict the search space and
improve the effectiveness of synthesis.  In our setting, types are
pairs of stochastic regular expressions, which come with a rich
algebraic structure that makes each regular expression semantically
equal to infinitely many others.  To tame that search space and
leverage the benefits of type-directed synthesis, we split our
algorithm into two communicating synthesizers following earlier
work~\cite{quotient-lens-synthesis}.  The first
synthesizer, \RXSearch, navigates the space of semantically equivalent
regular expressions by iteratively applying rewrite rules that
preserve both the semantics and the probability
distributions of those stochastic regular expressions. \RXSearch ranks
pairs of stochastic regular expressions 
by the number of rewrite rule applications required to obtain the pair
from the stochastic regular expressions given as input. It passes the
pairs off to the second synthesizer, \GreedySynth, in rank order.

At a high level, \GreedySynth looks for lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ that have the
best conditional entropy by performing a type-directed search. 
In more detail, \GreedySynth first uses type equalities based on semiring 
equivalences (ignoring star semiring equivalences) to normalize the
stochastic regular expressions provided by \RXSearch into 
\emph{stochastic n-ary DNF regular expressions}.  
Then, \GreedySynth uses the syntax of these n-ary DNF regular
expressions to find highly normalized lenses, in a
form we call \emph{simple symmetric n-ary DNF lenses}.  These lenses include
neither a composition operator nor a type equivalence rule.  These
restrictions mean that each simple symmetric n-ary DNF lens is typed
by a single pair of stochastic n-ary DNF regular expressions and they
make \GreedySynth's search space finite.  The variations that \GreedySynth
considers include choosing between \IdentityLens{} and  \Disconnect
lenses and deciding how to match multiple occurrences of the same kind
of data, \EG, the two occurrences of \lstinline{name} in the
management and HR data formats (we want to map first names to first
names and last names to last names). 
Finally, \GreedySynth converts this lens 
back into a simple symmetric lens by converting the n-ary forms into
the binary forms provided in the surface language.

As an example of the partnership between \RXSearch and \GreedySynth,
consider the lens shown in Figure~\ref{fig:example_lens}. 
This lens is \emph{not} in the set of lenses 
characterized by \lstinline{emp_salaries} and \lstinline{emp_insurance}, as that
typing derivation requires unrolling \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into \lstinline{"" | (("\n" . emp_salary) . ("\n" . emp_salary)$^*$)}.
